---
title: "Emergency Notifications for CED"
author: "Danylo Orlov, NYCC Data Team"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    keep_md: true  # This creates the intermediary .md file
  pdf_document:
    toc: true
  word_document:
    toc: true
---

```{r}
# Set global chunk options
knitr::opts_chunk$set(
  echo = TRUE,           # Show code chunks in output
  warning = FALSE,       # Don't show warnings
  message = FALSE,       # Don't show messages
  fig.width = 10,        # Default figure width
  fig.height = 6,        # Default figure height
  dpi = 300,             # Figure resolution
  fig.path = "figures/"  # Path to save figures
)
```

## Load Libraries

```{r}
library(readxl)
library(writexl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(stringr)
library(knitr)
library(Kendall)
library(kableExtra)
library(councilcount)
library(councildown)
library(councilverse)
```

## Uplaod File

```{r}
# Uplaod from Excel
emergency_data <- read_excel("~/Documents/Emergency_Notifs_for_CED/data/output/emergency_data_cleaned.xlsx")
```

## Analysis 1: Top Incident Types

```{r}
# Arrange in descending order the most common Incidents
incident_counts <- emergency_data %>%
  count(Incident, sort = TRUE) %>%
  rename(Count = n)

# Create a nice looking table
incident_counts %>%
  kable(col.names = c("Incident Type", "Number of Occurrences"),
        caption = "Frequency of Emergency Incident Types, All Districts") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2") %>%
  row_spec(1:3, background = "#f8f8f8", color = "red")
```

## Analysis 2: Top 3 Incident Types by District
### NOTE: "All Districts" refers to emergencies submitted to all Council Members.

```{r}
# Arrange in descending order top incident types by council district
top_incidents_by_district <- emergency_data %>%
  group_by(Council_District, Incident) %>%
  summarize(Count = n(), .groups = "drop") %>%
  group_by(Council_District) %>%
  arrange(desc(Count)) %>%
  slice_head(n = 3) %>%
  ungroup()

# Create a custom sorting function for districts
district_order <- function(x) {
  # Convert district to character if it's not already
  x <- as.character(x)
  # Create a numeric value for sorting
  ifelse(x == "All Districts", 100, # Put "All Districts" at the end
         ifelse(grepl("^\\d+$", x), as.numeric(x), 99)) # Convert numeric districts to numbers, others to 99
}

# Sort the data by district using the custom function
top_incidents_by_district <- top_incidents_by_district %>%
  mutate(sort_order = district_order(Council_District)) %>%
  arrange(sort_order) %>%
  select(-sort_order)

# Create a nice looking table without group_rows which is causing the error
top_incidents_by_district %>%
  kable(col.names = c("Council District", "Incident Type", "Frequency"),
        caption = "Top 3 Emergency Incident Types by Council District") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2") %>%
  # Add alternating background color for different districts
  row_spec(which(duplicated(top_incidents_by_district$Council_District) == FALSE), 
           extra_css = "border-top: 2px solid #666;") %>%
  footnote(general = "Note: Only top 3 incidents shown per district", 
           general_title = "",
           footnote_as_chunk = TRUE)
```

## Analysis 2.5: Total Incidents by District

```{r}
# Count total incidents by district
incidents_by_district <- emergency_data %>%
  group_by(Council_District) %>%
  summarize(Total_Incidents = n(), .groups = "drop")

# Create a custom sorting function for districts
district_order <- function(x) {
  # Convert district to character if it's not already
  x <- as.character(x)
  # Create a numeric value for sorting
  ifelse(x == "All Districts", 100, # Put "All Districts" at the end
         ifelse(grepl("^\\d+$", x), as.numeric(x), 99)) # Convert numeric districts to numbers, others to 99
}

# Sort the data by district using the custom function
incidents_by_district <- incidents_by_district %>%
  mutate(sort_order = district_order(Council_District)) %>%
  arrange(sort_order) %>%
  select(-sort_order)

# Create a nice looking table with minimal styling
incidents_by_district %>%
  kable(col.names = c("Council District", "Total Emergency Incidents"),
        caption = "Total Number of Emergency Incidents by Council District") %>%
  kable_styling(bootstrap_options = c("hover", "responsive"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2")
```

## Analysis 3: Top 3 Incident Types by Borough
### NOTE: "All Boroughs" refers to emergencies submitted to all Council Members.

```{r}
# Arrange in descending order top incident types by borough
top_incidents_by_borough <- emergency_data %>%
  group_by(Borough, Incident) %>%
  summarize(Count = n(), .groups = "drop") %>%
  group_by(Borough) %>%
  arrange(desc(Count)) %>%
  slice_head(n = 3) %>%
  ungroup() %>%
  arrange(Borough)

# Create a nice looking table without group_rows which is causing the error
top_incidents_by_borough %>%
  kable(col.names = c("Borough", "Incident Type", "Frequency"),
        caption = "Top 3 Emergency Incident Types by Borough") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2") %>%
  # Add alternating background color for different boroughs
  row_spec(which(duplicated(top_incidents_by_borough$Borough) == FALSE), 
           extra_css = "border-top: 2px solid #666;") %>%
  column_spec(3, bold = TRUE) %>%
  footnote(general = "Note: Only top 3 incidents shown per borough", 
           general_title = "",
           footnote_as_chunk = TRUE)
```

## Analysis 3.5: Total Incidents by Borough

```{r}
# Count total incidents by borough
incidents_by_borough <- emergency_data %>%
  group_by(Borough) %>%
  summarize(Total_Incidents = n(), .groups = "drop") %>%
  arrange(Borough)  # Simple alphabetical sort for boroughs

# Add percentage column
incidents_by_borough <- incidents_by_borough %>%
  mutate(Percentage = round(Total_Incidents / sum(Total_Incidents) * 100, 1))

# Create a clean table with grey header, including percentage column
incidents_by_borough %>%
  kable(col.names = c("Borough", "Total Emergency Incidents", "Percentage of All Incidents (%)"),
        caption = "Distribution of Emergency Incidents by Borough") %>%
  kable_styling(bootstrap_options = c("hover", "responsive"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2")
```

## Analysis 4: Trends 

```{r}
# Incidents per month 
incidents_by_month <- emergency_data %>%
  mutate(Month = floor_date(Date_of_Incident, "month")) %>%
  group_by(Month) %>%
  summarize(Total_Incidents = n(), .groups = "drop") %>%
  arrange(Month)

# Format months for better display
incidents_by_month <- incidents_by_month %>%
  mutate(Month_Formatted = format(Month, "%b %Y"))

# Create a nice looking table
incidents_by_month %>%
  select(Month_Formatted, Total_Incidents) %>%
  kable(col.names = c("Month", "Number of Emergency Incidents"),
        caption = "Monthly Trend of Emergency Incidents") %>%
  kable_styling(bootstrap_options = c("hover", "responsive"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2")

# Perform the Mann-Kendall test
mk_test <- MannKendall(incidents_by_month$Total_Incidents)
mk_p_value <- round(mk_test$sl[1], 4)  # p-value of the test
mk_tau <- round(mk_test$tau[1], 4)     # Kendall's tau statistic

# Determine trend direction and significance
trend_text <- ifelse(mk_p_value <= 0.05,
                     ifelse(mk_tau > 0, "Significant Increasing Trend", "Significant Decreasing Trend"),
                     "No Significant Trend")

# Make the plot
nycc_palette <- pal_nycc()
monthly_plot <- ggplot(incidents_by_month, aes(x = Month, y = Total_Incidents)) +
  geom_line(size = 0.5, color = nycc_palette[2]) +
  geom_point(size = 2, color = nycc_palette[2]) +
  # Add a smoother to highlight the trend visually
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", color = nycc_palette[4]) +
  labs(title = "Monthly Emergency Incidents",
       subtitle = paste0("Trend Analysis of Emergency Notifications\n",
                        "Mann-Kendall Test: ", trend_text, 
                        " (tau = ", mk_tau, ", p-value = ", mk_p_value, ")"),
       x = "Month",
       y = "Number of Incidents") +
  theme_nycc() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
print(monthly_plot)
```

## Analysis 4.5: Trends by Borough

```{r}
# Incidents per month by borough
incidents_by_month_borough <- emergency_data %>%
  mutate(Month = floor_date(Date_of_Incident, "month")) %>%
  group_by(Month, Borough) %>%
  summarize(Total_Incidents = n(), .groups = "drop") %>%
  arrange(Month, Borough)

# Format months for better display
incidents_by_month_borough <- incidents_by_month_borough %>%
  mutate(Month_Formatted = format(Month, "%b %Y"))

# Create a nice looking table
incidents_by_month_borough %>%
  select(Month_Formatted, Borough, Total_Incidents) %>%
  tidyr::pivot_wider(names_from = Borough, values_from = Total_Incidents) %>%
  kable(caption = "Monthly Emergency Incidents by Borough",
        align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = TRUE) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2")

# Get the list of unique boroughs
boroughs <- unique(incidents_by_month_borough$Borough)

# Create a list to store the plots
borough_plots <- list()

# Create individual plots for each borough with Mann-Kendall test
for (borough in boroughs) {
  # Filter data for the current borough
  borough_data <- incidents_by_month_borough %>%
    filter(Borough == borough)
  
  # Perform the Mann-Kendall test
  mk_test <- MannKendall(borough_data$Total_Incidents)
  mk_p_value <- round(mk_test$sl[1], 4)  # p-value of the test
  mk_tau <- round(mk_test$tau[1], 4)     # Kendall's tau statistic
  
  # Determine trend direction and significance
  trend_text <- ifelse(mk_p_value <= 0.05,
                       ifelse(mk_tau > 0, "Significant Increasing Trend", "Significant Decreasing Trend"),
                       "No Significant Trend")
  
  # Create the plot for this borough
  borough_plot <- ggplot(borough_data, aes(x = Month, y = Total_Incidents)) +
    geom_line(size = 0.5, color = nycc_palette[which(boroughs == borough)]) +
    geom_point(size = 2, color = nycc_palette[which(boroughs == borough)]) +
    # Add a smoother to highlight the trend visually
    geom_smooth(method = "loess", se = FALSE, linetype = "dashed", color = nycc_palette[4]) +
    labs(title = paste0(borough, " Emergency Incidents"),
         subtitle = paste0("Trend Analysis of Emergency Notifications\n",
                          "Mann-Kendall Test: ", trend_text, 
                          " (tau = ", mk_tau, ", p-value = ", mk_p_value, ")"),
         x = "Month",
         y = "Number of Incidents") +
    #theme_nycc() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
  
  # Store the plot in our list
  borough_plots[[borough]] <- borough_plot
  
  # Print the plot
  print(borough_plot)
}
```

## Analysis 5: Percentage of Concluded Reports by District

```{r}
# Percentage of concluded status incidents by council district
status_by_district <- emergency_data %>%
  group_by(Council_District) %>%
  summarize(
    Total_Incidents = n(),
    Concluded_Incidents = sum(Status == "Concluded"),
    Percentage_Concluded = round(Concluded_Incidents / Total_Incidents * 100, 1),
    .groups = "drop"
  )

# Create a custom sorting function for districts
district_order <- function(x) {
  # Convert district to character if it's not already
  x <- as.character(x)
  # Create a numeric value for sorting
  ifelse(x == "All Districts", 100, # Put "All Districts" at the end
         ifelse(grepl("^\\d+$", x), as.numeric(x), 99)) # Convert numeric districts to numbers, others to 99
}

# Sort the data by district using the custom function
status_by_district <- status_by_district %>%
  mutate(sort_order = district_order(Council_District)) %>%
  arrange(sort_order) %>%
  select(-sort_order)

# Create a simple table with grey header only, consistent with 2.5 and 3.5
status_by_district %>%
  kable(col.names = c("Council District", "Total Incidents", "Concluded Incidents", "% Concluded"),
        caption = "Resolution Status of Emergency Incidents by Council District") %>%
  kable_styling(bootstrap_options = c("hover", "responsive"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2")
```